Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-25 write-ups/FSE/paper.tex
Update overview figure
finished writing the bit about comparisons, added text around hugues figure

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{overview.pdf}
\caption{Cooperative kernels can flexibly be resized in order to let
other tasks (e.g. graphics) access GPU ressources}
\label{fig:overview}
\end{figure}

@@

For example, Nvidia's latest architecture (Pascal) is documented to
support instruction-level preemption~\cite{??}. Despite this, Nvidia's
most recent compute capablity does not guarantee forward progress for
CUDA programmers (evidence that vendors will not yet commit to providing such guarantees).
Indeed, on a GTX Titan X (Pascal) we still observe starvation:
we found that a global barrier can be executed successfully with 56 workgroups, but
deadlocks with 57 workgroups.  This is an example of a vendor
supporting preemption, yet not guaranteeing fair scheduling.

@@

@@

We present our cooperative kernels programming model as an extension
to OpenCL.  However, the cooperative kernels concept is more general,
and could be applied to extend other GPU programming models, such as
CUDA and HSA.  We describe the semantics of the model
(\mysec\ref{sec:semantics}), use our motivating examples to discuss
programmability (\mysec\ref{sec:programmingguidelines}), outline
important nonfunctional properties that the model requires to work
successfully (\mysec\ref{sec:nonfunctional}) and explain that the
model is backwards-compatible
(\mysec\ref{sec:backwardscompatibility}).
%
In Appendices~\ref{appendix:semantics} and~\ref{appendix:semanticalternatives} (supplementary material)
we present a formal semantics for cooperative kernels atop an abstract GPU programming model,
and discuss several cases where we
might have taken different and also reasonable semantic decisions.

@@

The number of \emph{active workgroups}---workgroups executing the
kernel---is denoted $M$.  Active workgroups
have consecutive ids in the range $[0, M-1]$.
Initially, at least one workgroup is active;
if necessary the scheduler must postpone the kernel until
some compute unit becomes available.

@@

Suppose a workgroup with id $m$ executes $\offerkill$.
If the workgroup has the highest largest id of all active workgroups then it can be killed by the scheduler, except that workgroup 0 can never be killed (to ensure that the kernel computation does not terminate early).  More formally, if $m < M-1$
or $M=1$ then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
M-1$, the scheduler can choose to ignore the offer, so that $\offerkill$
executes as a no-op, or accept the offer, so that execution of the workgroup ceases and
the number of active workgroups $M$ is atomically decremented by one.
%In the second case, the scheduler can assign some other task to
%the compute unit on which the workgroup was executing.

@@

@@

\mytablong\ref{tab:applications} gives an overview of the 8 irregular
algorithms that we ported to cooperative kernels. Among them, 6 are
graph algorithms, based on the Pannotia~\cite{Pannotia} and
Lonestar~\cite{BNP12} GPU application suites, using global barriers.
We indicate how many of the original number of barriers are changed to
resizing barriers, and how many variables need to be transmitted.  In
all cases all barriers were converted, except for bc which contains
barriers deeper in the call stack, a case not supported by our
prototype although these barriers could in principle be converted. The
remaining 2 algorithms are work stealing applications: each required
the addition of $\offerfork$ and $\offerkill$ at the start of the main
loop, and no variables needed to be transmitted.  For all these
examples, the porting to cooperative kernels was similarly
straightforward to the porting illustrated in
\mysec\ref{sec:programmingguidelines}. Most
graph come with 2 different data sets as input, leading to 13
application/input pairs in total.

From the Lonestar suite we omit two applications that use global
barriers: dmr has been shown to be unstable across GPU
vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially
short-running so that it is not interesting for our multitasking
approach.

@@

\begin{figure*}
\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf}
\includegraphics[width=.67\columnwidth]{hd520_mis_ecology.pdf}
\includegraphics[width=.67\columnwidth]{hd5500_bfs_rmat22.pdf}
%\vspace{-2mm}
\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
\vspace{-3mm}
\end{figure*}

\begin{table}
\normalsize
\centering
\begin{tabular}{ l l | l l | l l }
\multicolumn{2}{c|}{overall} & \multicolumn{2}{c|}{barrier} & \multicolumn{2}{c}{wk.steal.} \\
mean & max & mean & max & mean & max \\
\hline
$1.08$ & $1.37^{\ddagger}$ & $1.07$ & $1.37^{\ddagger}$ & $1.12$ & $1.23^{\diamond}$ \\
\end{tabular}\\
{\small
 $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
}
%\vspace{-2mm}
\caption{Cooperative kernel slowdown w/o multitasking}
\label{tab:overhead}
\vspace{-8mm}
\end{table}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{heavy.pdf}
%\vspace{-2mm}
\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
\vspace{-5mm}
\end{figure}

@@

Some recent GPUs, including NVIDIA's Pascal architecture, provide
hardware support for instruction-level preemption.  Few details of how
this works are available, however from reading NVIDIA documentation
and online material~\cite{ }, and running litmus tests on a Pascal
platform, our understanding is that preemption of entire kernels, but
not of individual workgroups running a kenrel is supported.
Intel---the only vendor whose OpenCL implemenation is capable enough
to implement cooperative kernels---do not provide this feature, thus
we cannot make a direct comparison.  We present here a theoretical
analysis of the overheads associated with sharing the GPU between
graphics and compute tasks via kernel-level preemption.

Let us assume that a graphics workload is required to be scheduled
with period $P$ and duration $D$, and that a long-running compute
kernel requires time $C$ to execute without interruption.  If we
assume the cost of preemption is negligible (e.g.\ NVIDIA have
reported preemption times of 0.1 ms for Pascal), then the overhead
associated with switching between compute and graphics every $P$ time
steps is $P/(P-D)$.

\ADComment{Tyler to complete.}

@@

\myparagraph{Preemptive multitasking on GPUs}
%
\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
%
Hardware support for preemption has been proposed for \nvidia GPUs,
as well as \emph{SM-draining}
whereby workgroups occupying a symmetric multiprocessor (SM; a compute
unit using our terminology) are allowed to complete until the SM becomes free for
other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited the presence of blocking constructs, since it may not be possible to drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing},
where a workgroup can be re-scheduled
from scratch if it has not yet committed side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
using simulators, over sets of regular GPU kernels.  Very recent \nvidia GPUs support
preemption, though it is not clear whether they guarantee fairness~\cite{PascalWhitepaper}.

The advantages of preemption are that it does not require programmer
effort, and guards against a rogue long-running kernel making the GPU
unavailable for other tasks.  The main drawback is that \emph{every}
kernel takes a potential performance hit.
%
If implemented natively, our solution does not impact performance on regular kernels that are not scheduled with cooperative kernels.
Furthermore, careful placement of
our new primitives can exploit problem structure, enabling more efficient multitasking than
application-oblivious preemption.  There is scope for combining the approaches: if a
cooperative kernel is insufficiently cooperative, hardware support for preemption could
be invoked to prevent the kernel from denying other tasks access to
the GPU indefinitely.

@@

\myparagraph{Optimizing GPU scheduling}
%
A number of works have considered how best to schedule dynamic
workloads on
GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.
Among these, the Whippletree approach employs a persistent megakernel
to schedule multiple, dynamic tasks in a manner that aims to best
utilise GPU resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}, and
the TimeGraph approach similarly aims to optimise scheduling of
competing workloads, in the context of
OpenGL~\cite{DBLP:conf/usenix/KatoLRI11}.  None of these approaches
tackle the problem of \emph{fair} scheduling on GPUs, thus they do not
aid in safe deployment of blocking irregular algorithms.

>>>>

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{overview.pdf}
\caption{Cooperative kernels can flexibly resize to let run other tasks
(e.g. graphics) concurrently}
\label{fig:overview}
\end{figure}

@@

For example, Nvidia's latest architecture (Pascal) is documented to
support instruction-level preemption~\cite{PascalWhitepaper}. Despite
this, Nvidia's most recent compute capablity does not guarantee
forward progress for CUDA programmers (evidence that vendors will not
yet commit to providing such guarantees).  Indeed, on a GTX Titan X
(Pascal) we still observe starvation: we found that a global barrier
can be executed successfully with 56 workgroups, but deadlocks with 57
workgroups.  This is an example of a vendor supporting preemption, yet
not guaranteeing fair scheduling.

@@

\myfiglong{\ref{fig:overview}} shows an example execution of how the
GPU compute units might be shared between a cooperative kernel and a
graphics task. Notice that workgroups executing the cooperative kernel
are killed at a $\offerkill$ command to make room for a graphics
task. After the graphics task, the previously killed workgroups are
restored to executing the cooperative kernel when a workgroup
executing the cooperative kernel calls $\offerfork$. The order in
which workgroups are killed is important and discussed in
\mysec{\ref{sec:semantics}}. The \emph{gather time} is the time
between when resources are requested and when the application has
surrendered the resources via $\offerkill$. To satisfy soft-real time
constraints, this time should be low; our experimental study
(\mysec{\ref{sec:responsiveness}}) shows that in practice, the
gather-time for our applications is acceptable for a range of graphics
workloads.

@@

We present our cooperative kernels programming model as an extension
to OpenCL.  However, the cooperative kernels concept is more general,
and could be applied to extend other GPU programming models, such as
CUDA and HSA.  We describe the semantics of the model
(\mysec\ref{sec:semantics}), use our motivating examples to discuss
programmability (\mysec\ref{sec:programmingguidelines}) and outline
important nonfunctional properties that the model requires to work
successfully (\mysec\ref{sec:nonfunctional}).
%and explain that the
%model is backwards-compatible
%(\mysec\ref{sec:backwardscompatibility}).
%
In Supplementary material we present a formal semantics for
cooperative kernels atop an abstract GPU programming model, and
discuss several cases where we might have taken different and also
reasonable semantic decisions.

@@

The number of \emph{active workgroups}---workgroups executing the
kernel---is denoted $M$.  Active workgroups have consecutive ids in
the range $[0, M-1]$.  Initially, at least one workgroup is active; if
necessary the scheduler must postpone the kernel until some compute
unit becomes available. For example, in \myfig{\ref{fig:overview}: at the
  beginning of the execution $M = 4$; while the graphics task is
  executing $M = 2$; after the fork $M = 4$ again.

@@

Suppose a workgroup with id $m$ executes $\offerkill$.  If the
workgroup has the highest largest id of all active workgroups then it
can be killed by the scheduler, except that workgroup 0 can never be
killed (to ensure that the kernel computation does not terminate
early).  More formally, if $m < M-1$ or $M=1$ then $\offerkill$ is a
no-op.  If instead $M > 1$ and $m = M-1$, the scheduler can choose to
ignore the offer, so that $\offerkill$ executes as a no-op, or accept
the offer, so that execution of the workgroup ceases and the number of
active workgroups $M$ is atomically decremented by one.
\myfiglong{\ref{fig:overview}} illustrates this, showing that
workgroup $3$ is killed before workgroup $2$.
%In the second case, the scheduler can assign some other task to
%the compute unit on which the workgroup was executing.

@@

\myfiglong{\ref{fig:overview}} illustrates the behaviour of
$\offerfork$. After the graphics task finishes executing, there is
space for two workgroups to join the computation of the cooperative
kernel. Workgroup $0$ calls $\offerfork$, spawning the two new
workgroups with ids $2$ and $3$ and broadcasting the transmit
variables to them. Workgroups $2$ and $3$ join the computation where
workgroup $0$ called $\offerfork$.

@@

\mytablong\ref{tab:applications} gives an overview of the 7 irregular
algorithms that we ported to cooperative kernels. Among them, 5 are
graph algorithms, based on the Pannotia~\cite{Pannotia} and
Lonestar~\cite{BNP12} GPU application suites, using global barriers.
We indicate how many of the original number of barriers are changed to
resizing barriers, and how many variables need to be transmitted.  In
all cases all barriers were converted. The remaining 2 algorithms are
work stealing applications: each required the addition of $\offerfork$
and $\offerkill$ at the start of the main loop, and no variables
needed to be transmitted.  For all these examples, the porting to
cooperative kernels was similarly straightforward to the porting
illustrated in \mysec\ref{sec:programmingguidelines}. Most graph come
with 2 different data sets as input, leading to 11 application/input
pairs in total.

%From the Lonestar suite we omit two applications that use global
%barriers: dmr has been shown to be unstable across GPU
%vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially
%short-running so that it is not interesting for our multitasking
%approach.

@@

\begin{figure*}
\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf}
\includegraphics[width=.67\columnwidth]{hd520_mis_ecology.pdf}
\includegraphics[width=.67\columnwidth]{hd5500_bfs_rmat22.pdf}
%\vspace{-2mm}
\caption{Example gather time and non-cooperative timing results \TSComment{need to get updated numbers}}\label{fig:fine-grained-timing}
\vspace{-3mm}
\end{figure*}

\begin{table}
\normalsize
\centering
\begin{tabular}{ l l | l l | l l }
\multicolumn{2}{c|}{overall} & \multicolumn{2}{c|}{barrier} & \multicolumn{2}{c}{wk.steal.} \\
mean & max & mean & max & mean & max \\
\hline
$1.09$ & $1.23^{\ddagger}$ & $1.09$ & $1.21^{\diamond}$ & $1.11$ & $1.23^{\ddagger}$ \\
\end{tabular}\\
{\small
 $^{\ddagger}$octree, $^{\diamond}$sssp usa\_ny
}
%\vspace{-2mm}
\caption{Cooperative kernel slowdown w/o multitasking \TSComment{need to get updated numbers}}
\label{tab:overhead}
\vspace{-8mm}
\end{table}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{heavy.pdf}
%\vspace{-2mm}
\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute \TSComment{need to get updated numbers}}\label{fig:performance}
\vspace{-5mm}
\end{figure}

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r}
g. workload &  kernel-level  & cooperative (resources) \\
\hline
light & 1.04 & 1.XX($N/4$)\\
medium & 1.08 & 1.XX($N/4$)\\
heavy & 1.33 & 1.XX($N-1$)\\
\end{tabular} \\
%\vspace{.2cm}

\caption{Overhead of kernel level preemption vs cooperative kernels for three
graphics workloads \TSComment{need to get updated numbers}}
\label{tab:preemption}
\vspace{-8mm}
\end{table}

Some recent GPUs, including NVIDIA's Pascal architecture, provide
hardware support for instruction-level preemption.  Few details of how
this works are available, however from reading NVIDIA documentation
and online material~\cite{PascalWhitepaper,anandtech}, and running
litmus tests on a Pascal platform, our understanding is that
preemption of entire kernels, but not of individual workgroups running
a kenrel is supported.  Intel---the only vendor whose OpenCL
implemenation is capable enough to implement cooperative kernels---do
not provide this feature, thus we cannot make a direct comparison.  We
present here a theoretical analysis of the overheads associated with
sharing the GPU between graphics and compute tasks via kernel-level
preemption.

Let us assume that a graphics workload is required to be scheduled
with period $P$ and duration $D$, and that a long-running compute
kernel requires time $C$ to execute without interruption.  If we
assume the cost of preemption is negligible (e.g.\ NVIDIA have
reported preemption times of 0.1 ms for Pascal), then the overhead
associated with switching between compute and graphics every $P$ time
steps is $P/(P-D)$. We compare this task-level preemption overhead
model with our experimental results per graphics workload in
\mytab{\ref{tab:preemption}}.

We see that our approach provides lower overhead for low and medium
graphics workloads. Recall for the heavy graphics workload, we are
within $10\%$ of the required deadline ($40$ ms) by using half the
workgroups with an overhead of $1.XX$, which is less than the
kernel-level preemption overhead of $1.33$. \TSComment{need to get
  updated numbers}

The low overhead of cooperative kernels for low and medium graphics
workloads are due to the irregularity of the cooperative kernels and
the property that they do not scale linearly. Thus some resources can
be taken away during computation without proportionally reducing
execution time.  Our poor performance for the heavy workload is
because the graphics task requires all but one workgroup, which
cripples the cooperative kernel enough that $\offerfork$ calls are not
issued as frequently.

These results suggest that a hybrid preemption scheme may work the
best for multi-tasking and graphics. That is, the cooperative approach
works well for light and medium tasks, but heavy graphics tasks
benefit from the coarser grained, kernel-level preemption strategy.

@@

\myparagraph{GPU multitasking and scheduling}
%
%\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
%
Hardware support for preemption has been proposed for \nvidia GPUs, as
well as \emph{SM-draining} whereby workgroups occupying a symmetric
multiprocessor (SM; a compute unit using our terminology) are allowed
to complete until the SM becomes free for other
tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited
the presence of blocking constructs, since it may not be possible to
drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing}, where a
workgroup can be re-scheduled from scratch if it has not yet committed
side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have
been evaluated using simulators, over sets of regular GPU kernels.
Very recent \nvidia GPUs (i.e. the Pascal architecture) support
preemption, though, as discussted in \mysec{sec:intro} and
\mysec{sec:nvidiacomparison}, it is not clear whether they guarantee
fairness or allow tasks to share GPU resources at the workgroup
level~\cite{PascalWhitepaper}.

%The advantages of preemption are that it does not require programmer
%effort, and guards against a rogue long-running kernel making the GPU
%unavailable for other tasks.  The main drawback is that \emph{every}
%kernel takes a potential performance hit.
%
%If implemented natively, our solution does not impact performance on
%regular kernels that are not scheduled with cooperative kernels.
%Furthermore, careful placement of our new primitives can exploit
%problem structure, enabling more efficient multitasking than
%application-oblivious preemption.  There is scope for combining the
%approaches: if a cooperative kernel is insufficiently cooperative,
%hardware support for preemption could be invoked to prevent the kernel
%from denying other tasks access to the GPU indefinitely.

@@

%\myparagraph{Optimizing GPU scheduling}
%
%A number of works have considered how best to schedule dynamic
%workloads on
%GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.
%Among these, the Whippletree approach employs a persistent megakernel
%to schedule multiple, dynamic tasks in a manner that aims to best
%utilise GPU resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}, and
%the TimeGraph approach similarly aims to optimise scheduling of
%competing workloads, in the context of
%OpenGL~\cite{DBLP:conf/usenix/KatoLRI11}.  None of these approaches
%tackle the problem of \emph{fair} scheduling on GPUs, thus they do not
%aid in safe deployment of blocking irregular algorithms.
